Structured Versus Decorated Cospans

### 2020-11-13 cospans.tex
More polishing of Theorem 2.1

Next we define the composite functor $\odot \colon F\lCsp_1 \times_{F\lCsp_0} F\lCsp_1 \to F\lCsp_1$, which amounts to defining how we compose two composable $F$-decorated cospans or two maps of $F$-decorated cospans with a common source and target. Given two composable $F$-decorated cospans which we respectively denote by $M$ and $N$, the capitals of their apices:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1,0) {$m$};
\node (C) at (2,0) {$b$};
\node (D) at (1,-0.5) {$x \in F(m)$};
\node (E) at (3,0) {$b$};
\node (F) at (4,0) {$n$};
\node (G) at (5,0) {$c$};
\node (H) at (4,-0.5) {$y \in F(n)$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$i$} (B)
(C) edge node[above]{$o$} (B)
(E) edge node[above]{$i'$} (F)
(G) edge node[above]{$o'$} (F);
\end{tikzpicture}
\]
the composite $M \odot N$ is given by taking the pushout of the two cospans in $\A$ over the shared foot which is the object $b$:
\begin{equation}\label{eq:dcospanscomposition}
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0) {$a$};
\node (B) at (1,1) {$m$};
\node (C) at (2,0) {$b$};
\node (D) at (3,1) {$n$};
\node (E) at (4,0) {$c$};
\node (F) at (2,2) {$m+n$};
\node (G) at (2,3) {$m+_{b} n$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$i$} (B)
(C) edge node[above]{$o$} (B)
(C) edge node [above] {$i'$} (D)
(E) edge node [above] {$o'$} (D)
(B) edge node [above] {$j$} (F)
(D) edge node [above] {$j'$} (F)
(F) edge node [left] {$\psi$} (G)
(A) edge[bend left] node [left] {$\psi j i$} (G)
(E) edge[bend right] node [right] {$\psi j' o'$} (G);
\end{tikzpicture}
\end{equation}
The composite decoration $x \odot y$ on the apex $m+_b n$ is given by:
$$x \odot y = 1 \xrightarrow{\lambda^{-1}} 1 \times 1 \xrightarrow{x \times y} F(m) \times F(n) \xrightarrow{\phi_{m,n}} F(m+n) \xrightarrow{F(\psi)} F(m+_{b}n)$$
where $\psi \maps m + n \to m+_{b} n$ is the natural map from the coproduct to the pushout and $\phi_{m,n} \maps F(m) \times F(n) \to F(m+n)$ is the natural transformation coming from the structure of the lax monoidal pseudofunctor $F \maps \A \to \Cat$. It is clear that the equations $S(M \odot N)=S(M)$ and $T(M \odot N)=T(N)$ hold. Given two horizontally composable maps of $F$-decorated cospans which we respectively denote by $\alpha$ and $\beta$:
\[
\begin{tikzpicture}[scale=1.5]
\node (A) at (0,0.5) {$a$};
\node (A') at (0,-0.5) {$a'$};
\node (B) at (1,0.5) {$m$};
\node (C) at (2,0.5) {$b$};
\node (C') at (2,-0.5) {$b'$};
\node (D) at (1,-0.5) {$m'$};
\node (E) at (3,0.5) {$x \in F(m)$};
\node (F) at (3,-0.5) {$x' \in F(m')$};
\node (G) at (4,0.5) {$b$};
\node (H) at (5,0.5) {$n$};
\node (I) at (6,0.5) {$c$};
\node (G') at (4,-0.5) {$b'$};
\node (H') at (5,-0.5) {$n'$};
\node (I') at (6,-0.5) {$c'$};
\node (J) at (7,0.5) {$y \in F(n)$};
\node (K) at (7,-0.5) {$y' \in F(n')$};
\node (L) at (1,-1) {$\tau_\alpha \maps F(h_1)(x) \to x'$};
\node (M) at (5,-1) {$\tau_\beta \maps F(h_2)(y) \to y'$};
\path[->,font=\scriptsize,>=angle 90]
(A) edge node[above]{$i_1$} (B)
(C) edge node[above]{$o_1$} (B)
(A) edge node[left]{$f$} (A')
(C) edge node[right]{$g$} (C')
(A') edge node[above] {$i_1'$} (D)
(C') edge node[above] {$o_1'$} (D)
(B) edge node [left] {$h_1$} (D)
(G) edge node [above] {$i_2$} (H)
(G) edge node [left] {$g$} (G')
(H) edge node [left] {$h_2$} (H')
(G') edge node [above] {$i_2'$} (H')
(I) edge node [above] {$o_2$} (H)
(I) edge node [right] {$k$} (I')
(I') edge node [above] {$o_2'$} (H');
\end{tikzpicture}
\]
\textbf{In progress...}

Given three composable horizontal 1-cells $M_1, M_2$ and $M_3$:

@@

\node (D) at (1,-0.5) {$x \in F(m_1)$};

@@

we get a natural isomorphism $a_{M_1,M_2,M_3} \maps (M_1 \odot M_2) \odot M_3 \to M_1 \odot (M_2 \odot M_3)$ which is a globular 2-morphism given by a map of cospans $(\id_{a_1},\sigma,\id_{a_4})$:

@@

\node (E) at (5.5,0.5) {$(x \odot y) \odot z \in F((m_1+_{b} m_2)+_{c} m_3)$};
\node (F) at (5.5,-0.5) {$x \odot (y \odot z) \in F(m_1+_{b} (m_2 +_{c} m_3))$};

@@

(B) edge node [left] {$\sigma$} (D);
with the decorations on the cospan's apices given by:
$$ (x \odot y) \odot z \mapseqq 1 \xrightarrow{\zeta_1} F(m_1+_{b} m_2) \times F(m_3) \xrightarrow{\phi_{m_1+_{b} m_2, m_3}} F((m_1+_{b}m_2) +m_3) \xrightarrow{F(j_{m_1+_{b} m_2,m_3})} F((m_1+_{b} m_2)+_{c} m_3)$$ $$\zeta_1 = (1 \times z) \rho^{-1} F(j_{m_1,m_2}) \phi_{m_1,m_2} (x \times y) \lambda^{-1}$$
$$ x \odot (y \odot z) \mapseqq 1 \xrightarrow{\zeta_2} F(m_1) \times F(m_2 +_{c} m_3) \xrightarrow{\phi_{m_1, m_2 +_{c} m_3}} F(m_1+(m_2 +_{c} m_3)) \xrightarrow{F(j_{m_1,m_2 +_{c} m_3})} F(m_1+_{b} (m_2+_{c} m_3))$$ $$\zeta_2 = (x \times 1) \lambda^{-1} F(j_{m_2,m_3}) \phi_{m_2,m_3} (y \odot z) \lambda^{-1}$$
together with the isomorphism $\tau_\sigma \maps F(\sigma)((x \odot y) \odot z) \to x \odot (y \odot z)$. Note that the map $\sigma \maps (m_1+_{b} m_2)+_{c}m_3 \to m_1+_{b}(m_2+_{c}m_3)$ is the universal map between two colimits of the same diagram. We also have left and right unitors where given a horizontal 1-cell $M$:

@@

To obtain the morphism of decorations for a horizontal composite, we have as initial data:
\node (A) at (4,0) {$\tau_\alpha \Swarrow$};
\node (E) at (4.5,0.5) {$F(m)$};
\node (E') at (4.5,-0.5) {$F(m')$};
\node (A') at (6.5,0) {$\tau_\beta \Swarrow$};
\node (D') at (5.5,0) {$1$};
\node (E'') at (7,0.5) {$F(n)$};
\node (E''') at (7,-0.5) {$F(n')$};
(D) edge node [above] {$x$} (E)
(D) edge node [below] {$x'$} (E')
(E) edge node [right] {$F(h_1)$} (E')
(D') edge node [above] {$y$} (E'')
(D') edge node [below] {$y'$} (E''')
(E'') edge node [right] {$F(h_2)$} (E''');
These two 2-morphisms $\tau_\alpha$ and $\tau_\beta$ are two 2-morphisms in the monoidal 2-category $(\Cat,\times,1)$ and so we can tensor them which results in:
(B) edge node [above] {$F(j_{m,n})$} (C)
(B') edge node [above] {$F(j_{m',n'})$} (C')
where the middle square commutes since $F$ is a lax monoidal pseudofunctor and the right square commutes as the underlying diagram commutes. The decorations $x \odot y$ and $x' \odot y'$ are given respectively by top and bottom composite of arrows and the morphism of decorations $\tau_{\alpha \odot \beta}$ is given by composing $\tau_\alpha \times \tau_\beta$ with the two commuting squares, which can equivalently be viewed as a morphism in $F(m' +_{b'} n')$.

>>>>

Next we define the composite functor $\odot \colon F\lCsp_1 \times_{F\lCsp_0} F\lCsp_1 \to F\lCsp_1$, which amounts to defining how we compose two composable $F$-decorated cospans or two maps of $F$-decorated cospans with a common source and target. Given two composable $F$-decorated cospans which we respectively denote by $M$ and $N$, the capitals of their apices:
\node (F) at (4,0) {$n$};
\node (H) at (4,-0.5) {$y \in F(n)$};
(G) edge node[above]{$o'$} (F);
the composite $M \odot N$ is given by taking the pushout of the two cospans in $\A$ over the shared foot which is the object $b$:
\begin{equation}\label{eq:dcospanscomposition}
\node (B) at (1,1) {$m$};
\node (D) at (3,1) {$n$};
\node (E) at (4,0) {$c$};
\node (F) at (2,2) {$m+n$};
\node (G) at (2,3) {$m+_{b} n$};
(C) edge node [above] {$i'$} (D)
(E) edge node [above] {$o'$} (D)
(B) edge node [above] {$j_m$} (F)
(D) edge node [above] {$j_{n}'$} (F)
(F) edge node [left] {$\psi$} (G)
(A) edge[bend left] node [left] {$\psi j_m i$} (G)
(E) edge[bend right] node [right] {$\psi j_{n}' o'$} (G);
The composite decoration $x \odot y$ on the apex $m+_b n$ is given by:
$$x \odot y = 1 \xrightarrow{\lambda^{-1}} 1 \times 1 \xrightarrow{x \times y} F(m) \times F(n) \xrightarrow{\phi_{m,n}} F(m+n) \xrightarrow{F(\psi)} F(m+_{b}n)$$
where $\psi \maps m + n \to m+_{b} n$ is the natural map from the coproduct to the pushout and $\phi_{m,n} \maps F(m) \times F(n) \to F(m+n)$ is the natural transformation coming from the structure of the lax monoidal pseudofunctor $F \maps \A \to \Cat$. Given two horizontally composable maps of $F$-decorated cospans which we respectively denote by $\alpha$ and $\beta$:
(I') edge node [above] {$o_2'$} (H');
their horizontal composite $\alpha \odot \beta$ is the pair given by first horizontally composing the two maps of cospans in $\A$:
\node (D) at (1.5,-0.5) {$m'+_{b'}n'$};
\node (E) at (4.5,0.5) {$x \odot y \in F(m+_b n)$};
(A) edge node[above]{$\psi j_m i_1$} (B)
(C) edge node[above]{$\psi j_n o_2$} (B)
(A') edge node [above]{$\psi j_{m'} i_1'$} (D)
(C') edge node [above]{$\psi j_{n'} o_2'$} (D)
(B) edge node [left] {$h_1 +_g h_2$} (D);
together with the morphism of decorations $\tau_{\alpha \odot \beta} \colon F(h_1 +_g h_2)(x \odot y) \to (x' \odot y')$ given by the following commutative diagram:
(B) edge node [above] {$F(\psi)$} (C)
(B') edge node [above] {$F(\psi)$} (C')
The middle square commutes since $F$ is a lax monoidal pseudofunctor and the right square commutes as the underlying square commutes. The decorations $x \odot y$ and $x' \odot y'$ are given respectively by the top and bottom composite of arrows and the morphism of decorations $\tau_{\alpha \odot \beta}$ is given by composing $\tau_\alpha \times \tau_\beta$ with the two commuting squares. It is clear that the equations $S(M \odot N)=S(M)$ and $T(M \odot N)=T(N)$ hold, and this finishes defining the four structure functors $U,S,T$ and $\odot$.

Lastly, our double category is required to have an associator $\alpha$ (\textbf{over using alpha, ignoring for now}) and left and right unitors $\lambda$ and $\rho$ which are natural isomorphisms governing the composition of horizontal 1-cells that are required to satisfy the standard coherence axioms for a monoidal category, as well as $S(\alpha),S(\lambda),S(\rho),T(\alpha),T(\lambda)$ and $T(\rho)$ all being identities. We will only show the associator $\alpha$, left unitor $\lambda$ and the pentagon identity, as the remaining details will be similar to these.

First we will build the associator $\alpha \colon \odot (\odot \times 1) \Rightarrow \odot (1 \times \odot)$, which will be a globular 2-morphism; in other words, a map of $F$-decorated cospans with identities on the sides. Given three composable horizontal 1-cells $M_1, M_2$ and $M_3$:

@@

@@

we first compute $(M_1 \odot M_2) \odot M_3)$ by pushing out the first two cospans followed by the third:
\node (B) at (2,0) {$(m_1 +_b m_2) +_c m_3$};
\node (C) at (4,0) {$b$};
%\node (D) at (3,0) {$x \in F(m)$};
(C) edge node[above]{$$} (B);
as well as the decoration $(x_1 \odot x_2) \odot x_3$ on the apex obtained as explained above. Explicitly, the decoration $(x_1 \odot x_2) \odot x_3) \in F((m_1+_b m_2) +_c m_3)$ is obtained as the composite $$ (x_1 \odot x_2) \odot x_3 \mapseqq 1 \xrightarrow{\zeta_1} F(m_1+_{b} m_2) \times F(m_3) \xrightarrow{\phi_{m_1+_{b} m_2, m_3}} F((m_1+_{b}m_2) +m_3) \xrightarrow{F(\psi)} F((m_1+_{b} m_2)+_{c} m_3)$$where$$\scriptstyle{\zeta_1 \mapseqq 1 \xrightarrow{\lambda^{-1}} 1 \times 1 \xrightarrow{x_1 \times x_2} F(m_1) \times F(m_2) \xrightarrow{\phi_{m_1,m_2}} F(m_1+m_2) \xrightarrow{F(j)} F(m_1+_b m_2) \xrightarrow {\rho^{-1}} F(m_1+_b m_2) \times 1 \xrightarrow{1 \times x_3} F(m_1+_b m_2) \times F(m_3).}$$
The composite $M_1 \odot (M_2 \odot M_3)$ is obtained similarly:
\node (B) at (2,0) {$m_1 +_b (m_2 +_c m_3)$};
\node (C) at (4,0) {$b$};
%\node (D) at (3,0) {$x \in F(m)$};
(C) edge node[above]{$$} (B);
as well as the decoration $x_1 \odot (x_2 \odot x_3)$ on the apex:
$$x_1 \odot (x_2 \odot x_3) \mapseqq 1 \xrightarrow{\zeta_2} F(m_1) \times F(m_2 +_c m_3) \xrightarrow{\phi_{m_1, m_2 +_c m_3}} F(m_1+(m_2 +_c +m_3)) \xrightarrow{F(\psi)} F(m_1+_{b} (m_2+_{c} m_3))$$with $\zeta_2$ being similar to $\zeta_1$ above. The associator $\alpha$ is then given by the map of $F$-decorated cospans given by the map of cospans

@@

\node (E) at (5.5,0.5) {$(x_1 \odot x_2) \odot x_3 \in F((m_1+_{b} m_2)+_{c} m_3)$};
\node (F) at (5.5,-0.5) {$x_1 \odot (x_2 \odot x_3) \in F(m_1+_{b} (m_2 +_{c} m_3))$};

@@

(B) edge node [left] {$\alpha$} (D);
together with the morphism of decorations $\tau_\alpha \colon F(\alpha)((x_1 \odot x_2) \odot x_3) \to (x_1 \odot (x_2 \odot x_3))$ that is obtained by pasting various commutative squares together, as we did for a horizontal composite above. The associator $\alpha$ will then satisfy the pentagon identity:
\node (A) at (-1,0) {$((M_1 \odot M_2) \odot M_3) \odot M_4$};
\node (B) at (3,1.05) {$(M_1 \odot (M_2 \odot M_3)) \odot M_4$};
\node (C) at (7,0) {$M_1 \odot ((M_2 \odot M_3) \odot M_4)$};
\node (D) at (1.5,-1.05) {$(M_1 \odot M_2) \odot (M_3 \odot M_4)$};
\node (E) at (4.5,-1.05) {$M_1 \odot (M_2 \odot (M_3 \odot M_4))$};
(A) edge node [above] {$\alpha \times 1$}(B)
(B) edge node [above] {$\alpha$}(C)
(A) edge node [below] {$\alpha$}(D)
(D) edge node [above] {$\alpha$}(E)
(C) edge node [below] {$1 \times \alpha$}(E);
which we may realize in two stages. First, for the underlying pentagon of maps of cospans in $\A$ commuting due to the fact that $\lCsp(\A)$ is a double category with its own associator that satisfies the pentagon identity, and second, for the pentagon of morphisms of decorations due to the pasting of various commutative squares in $\Cat$.

\noindent
\textbf{In progress...}
%we get a natural isomorphism $a_{M_1,M_2,M_3} \maps (M_1 \odot M_2) \odot M_3 \to M_1 \odot (M_2 \odot M_3)$ which is a globular 2-morphism given by a map of cospans $(\id_{a_1},\sigma,\id_{a_4})$:
%with the decorations on the cospan's apices given by:
%$$ (x \odot y) \odot z \mapseqq 1 \xrightarrow{\zeta_1} F(m_1+_{b} m_2) \times F(m_3) \xrightarrow{\phi_{m_1+_{b} m_2, m_3}} F((m_1+_{b}m_2) +m_3) \xrightarrow{F(j_{m_1+_{b} m_2,m_3})} F((m_1+_{b} m_2)+_{c} m_3)$$ $$\zeta_1 = (1 \times z) \rho^{-1} F(j_{m_1,m_2}) \phi_{m_1,m_2} (x \times y) \lambda^{-1}$$
%and
%$$ x \odot (y \odot z) \mapseqq 1 \xrightarrow{\zeta_2} F(m_1) \times F(m_2 +_{c} m_3) \xrightarrow{\phi_{m_1, m_2 +_{c} m_3}} F(m_1+(m_2 +_{c} m_3)) \xrightarrow{F(j_{m_1,m_2 +_{c} m_3})} F(m_1+_{b} (m_2+_{c} m_3))$$ $$\zeta_2 = (x \times 1) \lambda^{-1} F(j_{m_2,m_3}) \phi_{m_2,m_3} (y \odot z) \lambda^{-1}$$
%together with the isomorphism $\tau_\sigma \maps F(\sigma)((x \odot y) \odot z) \to x \odot (y \odot z)$. Note that the map $\sigma \maps (m_1+_{b} m_2)+_{c}m_3 \to m_1+_{b}(m_2+_{c}m_3)$ is the universal map between two colimits of the same diagram.
We also have left and right unitors where given a horizontal 1-cell $M$:

@@

%To obtain the morphism of decorations for a horizontal composite, we have as initial data:
%\node (A) at (4,0) {$\tau_\alpha \Swarrow$};
%\node (D) at (3,0) {$1$};
%\node (E) at (4.5,0.5) {$F(m)$};
%\node (E') at (4.5,-0.5) {$F(m')$};
%\node (A') at (6.5,0) {$\tau_\beta \Swarrow$};
%\node (D') at (5.5,0) {$1$};
%\node (E'') at (7,0.5) {$F(n)$};
%\node (E''') at (7,-0.5) {$F(n')$};
%(D) edge node [above] {$x$} (E)
%(D) edge node [below] {$x'$} (E')
%(E) edge node [right] {$F(h_1)$} (E')
%(D') edge node [above] {$y$} (E'')
%(D') edge node [below] {$y'$} (E''')
%(E'') edge node [right] {$F(h_2)$} (E''');
%These two 2-morphisms $\tau_\alpha$ and $\tau_\beta$ are two 2-morphisms in the monoidal 2-category $(\Cat,\times,1)$ and so we can tensor them which results in:
